---
title: "Hypoxic extent across cruises"
subtitle: "When and where dissolved oxygen falls below a hypoxic threshold (CTD casts)"
params:
  o2_threshold: 2.0
execute:
  echo: false
  message: false
  warning: false
---

This report summarizes **when** (by cruise date and depth from CTD casts) and **where** (from interpolated oxygen fields) dissolved oxygen falls below a hypoxic threshold. The default threshold is **2.0 mg/L**. Change `params.o2_threshold` in the YAML front matter or pass `--execute-params` when rendering:

```bash
quarto render reports/hypoxic_extent.qmd --execute-params '{"o2_threshold": 1.5}'
```

Run after processing and interpolation:

```bash
make process
make interpolate-all
make report-hypoxic
```

```{r setup}
ensure_packages <- function(pkgs) {
  for (pkg in pkgs) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg, repos = "https://cloud.r-project.org")
    }
  }
}

ensure_packages(c(
  "here", "dplyr", "ggplot2", "knitr", "scales", "viridisLite",
  "leaflet", "htmlwidgets", "jsonlite", "plotly", "terra", "sf", "patchwork"
))

library(here)
library(dplyr)
library(ggplot2)
library(knitr)
library(scales)
library(viridisLite)
library(leaflet)
library(htmlwidgets)
library(jsonlite)

o2_threshold <- params$o2_threshold
if (!is.numeric(o2_threshold) || length(o2_threshold) != 1 || !is.finite(o2_threshold)) {
  stop("params$o2_threshold must be a single finite number (mg/L).")
}

source(here("R/hypoxic_summary.R"))
source(here("R/hypoxic_field_summary.R"))
source(here("R/plot_hypoxic_frequency_map.R"))
source(here("R/volume_bathymetry.R"))
source(here("R/plot_hypoxic_volume.R"))

all_obs <- load_all_cruise_observations()
field_layers <- depth_layers_for_observations(all_obs)

cruise_summary <- summarize_all_cruises_hypoxic(
  o2_threshold,
  observations = all_obs
)
depth_extent <- hypoxic_extent_by_depth_all_cruises(
  o2_threshold,
  observations = all_obs
)
hypoxic_obs <- hypoxic_observations_all_cruises(
  o2_threshold,
  observations = all_obs
)
field_frequency <- hypoxic_field_frequency_all_cruises(
  o2_threshold,
  layers = field_layers
)
field_min_depth_by_cruise <- hypoxic_min_depth_by_cruise(
  o2_threshold
)
calendar_heatmap_df <- prepare_hypoxic_calendar_heatmap(field_min_depth_by_cruise)
volume_grid <- hypoxic_field_volume_frequency(o2_threshold, subsample = 2L)
n_cruises <- nrow(cruise_summary)
n_interp_cruises <- length(list_interpolated_cruises())
n_hypoxic_casts <- if (nrow(hypoxic_obs) > 0) nrow(hypoxic_obs) else 0L
```

## Overview

| Metric | Value |
|--------|------:|
| O₂ threshold (mg/L) | `r o2_threshold` |
| Cruises with cleaned CTD data | `r n_cruises` |
| Cruises with interpolated fields | `r n_interp_cruises` |
| Cruises with hypoxic casts | `r sum(cruise_summary$pct_casts_hypoxic > 0, na.rm = TRUE)` |
| Hypoxic casts | `r n_hypoxic_casts` |
| Total CTD observations | `r sum(cruise_summary$n_obs, na.rm = TRUE)` |

## When: hypoxic extent over time

The timeline below tracks hypoxia in each cruise's CTD casts. **Surface hypoxic** is the share of casts whose shallowest sample is below the threshold; **any depth** is the share of casts with at least one sample below the threshold. **Median cast depth** (right axis) is the median deepest sample depth per cruise.

```{r timeline}
#| fig-cap: "Hypoxic cast coverage by cruise date."

timeline_colors <- c(
  "Shallowest sample" = "#D55E00",
  "Any depth in cast" = "#0072B2",
  "Median cast depth" = "#009E73"
)

if (n_cruises == 0) {
  cat("No cleaned CTD data found. Run `make process` first.")
} else {
  timeline_df <- cruise_summary |> filter(!is.na(cruise_date))
  max_cast_depth <- max(timeline_df$median_cast_depth_m, na.rm = TRUE)

  ggplot(timeline_df, aes(x = cruise_date)) +
    geom_line(
      aes(y = pct_surface_hypoxic, color = "Shallowest sample"),
      linewidth = 0.85
    ) +
    geom_point(
      aes(y = pct_surface_hypoxic, color = "Shallowest sample"),
      size = 1.8
    ) +
    geom_line(
      aes(y = pct_casts_hypoxic, color = "Any depth in cast"),
      linewidth = 0.85
    ) +
    geom_point(
      aes(y = pct_casts_hypoxic, color = "Any depth in cast"),
      size = 1.8
    ) +
    geom_line(
      aes(y = median_cast_depth_m / max_cast_depth * 100, color = "Median cast depth"),
      linewidth = 0.85,
      linetype = "22"
    ) +
    geom_point(
      aes(y = median_cast_depth_m / max_cast_depth * 100, color = "Median cast depth"),
      size = 1.8
    ) +
    scale_color_manual(name = NULL, values = timeline_colors) +
    scale_y_continuous(
      limits = c(0, 100),
      labels = label_number(suffix = "%"),
      name = paste0("Casts below ", o2_threshold, " mg/L"),
      sec.axis = sec_axis(
        trans = ~ . * max_cast_depth / 100,
        name = "Median cast depth (m)",
        labels = label_number(accuracy = 1)
      )
    ) +
    labs(x = "Cruise date") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
}
```

Median shallow hypoxic depth summarizes how close to the surface hypoxia begins in each cruise (among casts that go hypoxic at any depth). **Median cast depth** (right axis) shows how deep casts typically sampled.

```{r median-depth-timeline}
#| fig-cap: "Median shallowest hypoxic depth among hypoxic casts, by cruise."

depth_timeline_colors <- c(
  "Median shallow hypoxic depth" = "#0072B2",
  "Median cast depth" = "#D55E00"
)

if (n_cruises == 0) {
  cat("No cleaned CTD data found.")
} else {
  depth_df <- cruise_summary |>
    filter(
      !is.na(cruise_date),
      is.finite(median_shallow_hypoxic_depth_m) | is.finite(median_cast_depth_m)
    )
  hyp_max <- max(depth_df$median_shallow_hypoxic_depth_m, na.rm = TRUE)
  cast_max <- max(depth_df$median_cast_depth_m, na.rm = TRUE)

  ggplot(depth_df, aes(x = cruise_date)) +
    geom_line(
      aes(y = median_shallow_hypoxic_depth_m, color = "Median shallow hypoxic depth"),
      linewidth = 0.85
    ) +
    geom_point(
      aes(y = median_shallow_hypoxic_depth_m, color = "Median shallow hypoxic depth"),
      size = 1.8
    ) +
    geom_line(
      aes(y = median_cast_depth_m * hyp_max / cast_max, color = "Median cast depth"),
      linewidth = 0.85,
      linetype = "22"
    ) +
    geom_point(
      aes(y = median_cast_depth_m * hyp_max / cast_max, color = "Median cast depth"),
      size = 1.8
    ) +
    scale_color_manual(name = NULL, values = depth_timeline_colors) +
    scale_y_reverse(
      name = "Median shallow hypoxic depth (m)",
      limits = c(hyp_max * 1.05, 0),
      sec.axis = sec_axis(
        trans = ~ . * cast_max / hyp_max,
        name = "Median cast depth (m)",
        labels = label_number(accuracy = 1)
      )
    ) +
    labs(x = "Cruise date") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
}
```

## When: calendar heatmap from interpolated fields

Each tile is a cruise week (ISO year × week). Color shows the **shallowest hypoxic depth (m)** in the interpolated oxygen field— the minimum depth at which O₂ falls below the threshold (default **2.0 mg/L**, set via `params.o2_threshold`). Shallow hypoxia is yellow; deeper hypoxia is purple. Weeks with no hypoxia are left blank.

```{r calendar-heatmap}
#| fig-cap: "Shallowest hypoxic depth by cruise week from interpolated fields."
#| fig-height: 10

if (n_interp_cruises == 0) {
  cat("No interpolated oxygen fields found. Run `make interpolate-all` first.")
} else if (nrow(calendar_heatmap_df) == 0) {
  cat("No hypoxic interpolated fields found below ", o2_threshold, " mg/L.", sep = "")
} else {
  plot_hypoxic_calendar_heatmap(calendar_heatmap_df, o2_threshold)
}
```

## When: hypoxic extent by depth

Each row is a cruise; color is the percentage of CTD observations below the threshold in each depth bin.

```{r depth-heatmap}
#| fig-cap: "Percent of CTD observations below the hypoxic threshold at each depth."
#| fig-height: 10

depth_bin_m <- 4

if (nrow(depth_extent) == 0) {
  cat("No depth-level summaries available.")
} else {
  heatmap_df <- prepare_hypoxic_depth_heatmap(depth_extent, bin_m = depth_bin_m) |>
    mutate(
      cruise_label = factor(
        paste0(cruise_id, " (", format(cruise_date, "%Y-%m"), ")"),
        levels = cruise_labels_ordered(depth_extent)
      )
    )

  ggplot(
    heatmap_df,
    aes(
      x = depth_bin + depth_bin_m / 2,
      y = cruise_label,
      fill = pct_hypoxic
    )
  ) +
    geom_tile(width = depth_bin_m, height = 0.85) +
    scale_fill_viridis_c(
      option = "C",
      limits = c(0, 100),
      name = "% hypoxic"
    ) +
    scale_x_continuous(expand = expansion(mult = c(0, 0.01))) +
    labs(
      x = "Depth (m)",
      y = NULL
    ) +
    theme_minimal(base_size = 11) +
    theme(axis.text.y = element_text(size = 7))
}
```

## Where: 3D hypoxic volume

Interactive 3D view of hypoxic regions from interpolated fields, with **BlueTopo bathymetry** as the seafloor surface and **coastline basemap lines** at the sea surface. Each voxel shows the share of cruises with O₂ below the threshold at that location and depth (aligned to a common 4 m depth grid). Hypoxic volumes render at **60% opacity**; non-hypoxic space is transparent. The color bar uses a **log scale (1–100%)**. Drag to rotate, scroll to zoom.

```{r volume-3d}
#| code-summary: Interactive 3D volumetric display of hypoxic regions

if (n_interp_cruises == 0) {
  cat("No interpolated oxygen fields found. Run `make interpolate-all` first.")
} else if (nrow(volume_grid) == 0) {
  cat("No hypoxic voxels found in interpolated fields below ", o2_threshold, " mg/L.", sep = "")
} else {
  volume <- plot_hypoxic_volume(
    volume_grid,
    o2_threshold,
    title = paste0(
      "Hypoxic volume across cruises (O\u2082 < ",
      o2_threshold,
      " mg/L)"
    )
  )
  if (is.null(volume)) {
    cat("Unable to build hypoxic volume display.")
  } else {
    volume
  }
}
```

## Where: hypoxia frequency from interpolated fields

Use **Prev** / **Next** to step through ecological depth layers (same bins as the [oxygen field maps](../cruise_oxygen/listing.qmd)). Color shows the **log-scaled percentage of cruises** whose interpolated field is hypoxic (O₂ below the threshold) at each grid cell within that layer.

```{r spatial-map}
#| code-summary: Interactive heatmap of hypoxia frequency from interpolated fields

if (n_interp_cruises == 0) {
  cat("No interpolated oxygen fields found. Run `make interpolate-all` first.")
} else if (nrow(field_frequency) == 0) {
  cat("No hypoxic cells found in interpolated fields below ", o2_threshold, " mg/L.", sep = "")
} else {
  map <- plot_hypoxic_frequency_map(field_frequency, field_layers, o2_threshold)
  if (is.null(map)) {
    cat("Unable to build hypoxia frequency map.")
  } else {
    map
  }
}
```

## Cruise summary table

```{r cruise-table}
if (n_cruises == 0) {
  cat("No cruises to summarize.")
} else {
  cruise_summary |>
    mutate(
      cruise_date = format(cruise_date, "%Y-%m-%d"),
      pct_casts_hypoxic = round(pct_casts_hypoxic, 1),
      pct_surface_hypoxic = round(pct_surface_hypoxic, 1),
      pct_obs_hypoxic = round(pct_obs_hypoxic, 1),
      median_shallow_hypoxic_depth_m = round(median_shallow_hypoxic_depth_m, 1),
      min_obs_o2_mg_l = round(min_obs_o2_mg_l, 2),
      max_obs_o2_mg_l = round(max_obs_o2_mg_l, 2)
    ) |>
    select(
      cruise_id, cruise_date, n_casts, n_hypoxic_casts, n_obs,
      pct_surface_hypoxic, pct_casts_hypoxic, pct_obs_hypoxic,
      median_shallow_hypoxic_depth_m, min_obs_o2_mg_l, max_obs_o2_mg_l
    ) |>
    kable(col.names = c(
      "Cruise", "Date", "Casts", "Hypoxic casts", "Observations",
      "% shallow hypoxic", "% casts hypoxic", "% obs hypoxic",
      "Median shallow hypoxic depth (m)", "Min O₂ (mg/L)", "Max O₂ (mg/L)"
    ))
}
```

Generated on `r format(Sys.time(), "%Y-%m-%d %H:%M %Z")`.
