---
title: "Hypoxic extent across cruises"
subtitle: "When and where dissolved oxygen falls below a hypoxic threshold (CTD casts)"
params:
  o2_threshold: 2.0
execute:
  echo: false
  message: false
  warning: false
---

This report summarizes **when** (by cruise date and depth) and **where** (across the study domain) cleaned **CTD cast** dissolved oxygen falls below a hypoxic threshold. The default threshold is **2.0 mg/L**. Change `params.o2_threshold` in the YAML front matter or pass `--execute-params` when rendering:

```bash
quarto render reports/hypoxic_extent.qmd --execute-params '{"o2_threshold": 1.5}'
```

Run after processing:

```bash
make process
make report-hypoxic
```

```{r setup}
ensure_packages <- function(pkgs) {
  for (pkg in pkgs) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg, repos = "https://cloud.r-project.org")
    }
  }
}

ensure_packages(c(
  "here", "dplyr", "ggplot2", "knitr", "leaflet", "scales", "viridisLite"
))

library(here)
library(dplyr)
library(ggplot2)
library(knitr)
library(leaflet)
library(scales)
library(viridisLite)

o2_threshold <- params$o2_threshold
if (!is.numeric(o2_threshold) || length(o2_threshold) != 1 || !is.finite(o2_threshold)) {
  stop("params$o2_threshold must be a single finite number (mg/L).")
}

source(here("R/hypoxic_summary.R"))

cruise_summary <- summarize_all_cruises_hypoxic(o2_threshold)
depth_extent <- hypoxic_extent_by_depth_all_cruises(o2_threshold)
hypoxic_obs <- hypoxic_observations_all_cruises(o2_threshold)
n_cruises <- nrow(cruise_summary)
n_hypoxic_casts <- if (nrow(hypoxic_obs) > 0) nrow(hypoxic_obs) else 0L
```

## Overview

| Metric | Value |
|--------|------:|
| O₂ threshold (mg/L) | `r o2_threshold` |
| Cruises with cleaned CTD data | `r n_cruises` |
| Cruises with hypoxic casts | `r sum(cruise_summary$pct_casts_hypoxic > 0, na.rm = TRUE)` |
| Hypoxic casts | `r n_hypoxic_casts` |
| Total CTD observations | `r sum(cruise_summary$n_obs, na.rm = TRUE)` |

## When: hypoxic extent over time

The timeline below tracks hypoxia in each cruise's CTD casts. **Surface hypoxic** is the share of casts whose shallowest sample is below the threshold; **any depth** is the share of casts with at least one sample below the threshold.

```{r timeline}
#| fig-cap: "Hypoxic cast coverage by cruise date."

if (n_cruises == 0) {
  cat("No cleaned CTD data found. Run `make process` first.")
} else {
  cruise_summary |>
    filter(!is.na(cruise_date)) |>
    ggplot(aes(x = cruise_date)) +
    geom_line(aes(y = pct_surface_hypoxic, color = "Shallowest sample"), linewidth = 0.7) +
    geom_point(aes(y = pct_surface_hypoxic, color = "Shallowest sample"), size = 1.8) +
    geom_line(aes(y = pct_casts_hypoxic, color = "Any depth in cast"), linewidth = 0.7) +
    geom_point(aes(y = pct_casts_hypoxic, color = "Any depth in cast"), size = 1.8) +
    scale_color_manual(
      name = NULL,
      values = c("Shallowest sample" = "#D95F0E", "Any depth in cast" = "#2C7FB8")
    ) +
    scale_y_continuous(limits = c(0, 100), labels = label_number(suffix = "%")) +
    labs(
      x = "Cruise date",
      y = paste0("Casts below ", o2_threshold, " mg/L")
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
}
```

Median shallow hypoxic depth summarizes how close to the surface hypoxia begins in each cruise (among casts that go hypoxic at any depth):

```{r median-depth-timeline}
#| fig-cap: "Median shallowest hypoxic depth among hypoxic casts, by cruise."

if (n_cruises == 0) {
  cat("No cleaned CTD data found.")
} else {
  ggplot(
    cruise_summary |> filter(!is.na(cruise_date), is.finite(median_shallow_hypoxic_depth_m)),
    aes(x = cruise_date, y = median_shallow_hypoxic_depth_m)
  ) +
    geom_line(color = "#756BB1", linewidth = 0.7) +
    geom_point(color = "#756BB1", size = 1.8) +
    scale_y_reverse() +
    labs(
      x = "Cruise date",
      y = "Median shallow hypoxic depth (m)"
    ) +
    theme_minimal(base_size = 12)
}
```

## When: hypoxic extent by depth

Each row is a cruise; color is the percentage of CTD observations below the threshold in each depth bin.

```{r depth-heatmap}
#| fig-cap: "Percent of CTD observations below the hypoxic threshold at each depth."
#| fig-height: 10

depth_bin_m <- 4

if (nrow(depth_extent) == 0) {
  cat("No depth-level summaries available.")
} else {
  depth_labels <- depth_extent |>
    filter(!is.na(cruise_date)) |>
    arrange(cruise_date, cruise_id) |>
    distinct(cruise_id, cruise_date) |>
    mutate(label = paste0(cruise_id, " (", format(cruise_date, "%Y-%m"), ")")) |>
    pull(label)

  heatmap_df <- prepare_hypoxic_depth_heatmap(depth_extent, bin_m = depth_bin_m) |>
    mutate(
      cruise_label = factor(
        paste0(cruise_id, " (", format(cruise_date, "%Y-%m"), ")"),
        levels = depth_labels
      )
    )

  ggplot(
    heatmap_df,
    aes(
      x = depth_bin + depth_bin_m / 2,
      y = cruise_label,
      fill = pct_hypoxic
    )
  ) +
    geom_tile(width = depth_bin_m, height = 0.85) +
    scale_fill_viridis_c(
      option = "C",
      limits = c(0, 100),
      name = "% hypoxic"
    ) +
    scale_x_continuous(expand = expansion(mult = c(0, 0.01))) +
    labs(
      x = "Depth (m)",
      y = NULL
    ) +
    theme_minimal(base_size = 11) +
    theme(axis.text.y = element_text(size = 7))
}
```

## Where: cruise domains and hypoxic casts

Cruise centroids (mean cast location) are colored by the share of casts with any hypoxic sample. Points show individual hypoxic casts.

```{r spatial-map}
#| fig-cap: "Cruise-level hypoxic extent and hypoxic CTD casts."

if (n_cruises == 0) {
  cat("No cleaned CTD data found.")
} else {
  pal <- colorNumeric(
    palette = viridisLite::viridis(256, option = "C"),
    domain = c(0, max(cruise_summary$pct_casts_hypoxic, na.rm = TRUE)),
    na.color = "#cccccc"
  )

  map <- leaflet(cruise_summary, width = "100%", height = 520) |>
    addProviderTiles(providers$Esri.OceanBasemap, group = "Ocean") |>
    addProviderTiles(providers$CartoDB.Positron, group = "Light") |>
    addCircleMarkers(
      lng = ~mean_lon,
      lat = ~mean_lat,
      radius = 7,
      stroke = TRUE,
      weight = 1,
      color = "#1a1a1a",
      fillColor = ~pal(pct_casts_hypoxic),
      fillOpacity = 0.85,
      label = ~sprintf(
        "%s\n%.0f%% casts hypoxic\nmedian shallow depth: %s m",
        cruise_id,
        pct_casts_hypoxic,
        ifelse(
          is.finite(median_shallow_hypoxic_depth_m),
          sprintf("%.1f", median_shallow_hypoxic_depth_m),
          "—"
        )
      )
    )

  if (nrow(hypoxic_obs) > 0) {
    map <- map |>
      addCircleMarkers(
        data = hypoxic_obs,
        lng = ~longitude,
        lat = ~latitude,
        radius = 4,
        color = "#b30000",
        fillColor = "#fc9272",
        fillOpacity = 0.75,
        weight = 1,
        label = ~sprintf(
          "%s / cast %s\nmin O\u2082: %.2f mg/L\nshallowest hypoxic depth: %.1f m",
          cruise_id,
          cast_id,
          min_o2_mg_l,
          shallowest_hypoxic_depth_m
        ),
        group = "Hypoxic casts"
      ) |>
      addLayersControl(
        baseGroups = c("Ocean", "Light"),
        overlayGroups = c("Hypoxic casts"),
        options = layersControlOptions(collapsed = TRUE)
      )
  } else {
    map <- map |>
      addLayersControl(
        baseGroups = c("Ocean", "Light"),
        options = layersControlOptions(collapsed = TRUE)
      )
  }

  map |>
    addLegend(
      pal = pal,
      values = ~pct_casts_hypoxic,
      title = paste0("% casts below ", o2_threshold, " mg/L"),
      position = "bottomright"
    ) |>
    fitBounds(
      lng1 = min(cruise_summary$mean_lon, na.rm = TRUE),
      lat1 = min(cruise_summary$mean_lat, na.rm = TRUE),
      lng2 = max(cruise_summary$mean_lon, na.rm = TRUE),
      lat2 = max(cruise_summary$mean_lat, na.rm = TRUE)
    )
}
```

## Cruise summary table

```{r cruise-table}
if (n_cruises == 0) {
  cat("No cruises to summarize.")
} else {
  cruise_summary |>
    mutate(
      cruise_date = format(cruise_date, "%Y-%m-%d"),
      pct_casts_hypoxic = round(pct_casts_hypoxic, 1),
      pct_surface_hypoxic = round(pct_surface_hypoxic, 1),
      pct_obs_hypoxic = round(pct_obs_hypoxic, 1),
      median_shallow_hypoxic_depth_m = round(median_shallow_hypoxic_depth_m, 1),
      min_obs_o2_mg_l = round(min_obs_o2_mg_l, 2),
      max_obs_o2_mg_l = round(max_obs_o2_mg_l, 2)
    ) |>
    select(
      cruise_id, cruise_date, n_casts, n_hypoxic_casts, n_obs,
      pct_surface_hypoxic, pct_casts_hypoxic, pct_obs_hypoxic,
      median_shallow_hypoxic_depth_m, min_obs_o2_mg_l, max_obs_o2_mg_l
    ) |>
    kable(col.names = c(
      "Cruise", "Date", "Casts", "Hypoxic casts", "Observations",
      "% shallow hypoxic", "% casts hypoxic", "% obs hypoxic",
      "Median shallow hypoxic depth (m)", "Min O₂ (mg/L)", "Max O₂ (mg/L)"
    ))
}
```

Generated on `r format(Sys.time(), "%Y-%m-%d %H:%M %Z")`.
