---
title: "CTD Processing Summary"
subtitle: "Changes applied during `make process`"
format:
  html:
    toc: true
    code-fold: true
    theme: cosmo
    css: ../styles.css
execute:
  echo: false
  message: false
  warning: false
---

This report summarizes how CTD cast data changed when raw ERDDAP downloads in `data/01_raw/` were cleaned into `data/02_clean/`.

Run after processing:

```bash
make process
make report-process
```

## Processing steps

Each cast file is processed with the [`oce`](https://dankelley.github.io/oce/) package using the same workflow as [`ctd_reports_template.qmd`](../ctd_reports_template.qmd):

1. **Load** — build an `oce` CTD object from the raw CSV (`R/ctd_load_from_csv.R`)
2. **Trim** — `oce::ctdTrim()` removes pre-deployment and post-recovery scans
3. **Decimate** — `oce::ctdDecimate(p = 0.1)` reduces vertical resolution to 0.1 dbar bins
4. **Export** — keep the original CSV columns for scans retained after cleaning; add `station` and `cruise_id`

Rows without `sea_water_pressure` are excluded before scan matching.

```{r setup}
ensure_packages <- function(pkgs) {
  for (pkg in pkgs) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg, repos = "https://cloud.r-project.org")
    }
  }
}

ensure_packages(c("here", "readr", "dplyr", "knitr", "ggplot2"))

library(here)
library(readr)
library(dplyr)
library(knitr)
library(ggplot2)

source(here("R/summarize_processing.R"))

cast_summary <- summarize_all_processing()
cruise_summary <- summarize_processing_by_cruise(cast_summary)

total_raw <- sum(cast_summary$raw_rows, na.rm = TRUE)
total_clean <- sum(cast_summary$clean_rows, na.rm = TRUE)
total_removed <- total_raw - total_clean
overall_pct_kept <- if (total_raw > 0) round(100 * total_clean / total_raw, 1) else NA_real_

status_counts <- cast_summary %>%
  count(status, name = "casts")
```

## Overall summary

| Metric | Value |
|--------|------:|
| Casts processed | `r nrow(cast_summary)` |
| Cruises | `r nrow(cruise_summary)` |
| Raw rows (with pressure) | `r format(total_raw, big.mark = ",")` |
| Clean rows | `r format(total_clean, big.mark = ",")` |
| Rows removed | `r format(total_removed, big.mark = ",")` |
| Overall percent kept | `r overall_pct_kept`% |

### Status by cast

```{r status-table}
kable(status_counts, col.names = c("Status", "Casts"))
```

## Changes by cruise

```{r cruise-table}
cruise_summary %>%
  arrange(desc(pct_kept)) %>%
  kable(digits = 1, col.names = c(
    "Cruise", "Casts", "Cleaned", "Skipped", "Empty", "Failed",
    "Raw rows", "Clean rows", "% kept"
  ))
```

```{r cruise-plot}
#| fig-cap: "Percent of raw rows retained after cleaning, by cruise."
ggplot(cruise_summary, aes(x = reorder(cruise_id, pct_kept), y = pct_kept)) +
  geom_col(fill = "#2C7FB8") +
  coord_flip() +
  labs(x = NULL, y = "Percent of raw rows kept") +
  theme_minimal(base_size = 12)
```

## Cast-level detail

Casts with the largest row reduction (potential outliers to review):

```{r cast-detail}
cast_summary %>%
  filter(status %in% c("cleaned", "skipped")) %>%
  arrange(desc(pct_removed)) %>%
  select(cruise_id, cast_id, station, status, raw_rows, clean_rows, rows_removed, pct_removed) %>%
  head(25) %>%
  kable(digits = 1)
```

Casts that failed or produced no output:

```{r cast-problems}
problems <- cast_summary %>%
  filter(status %in% c("failed", "empty", "missing_output"))

if (nrow(problems) == 0) {
  cat("No failed or empty casts.")
} else {
  problems %>%
    select(cruise_id, cast_id, status, raw_rows, clean_rows) %>%
    kable()
}
```

## Example profile comparison

The cast with the largest row reduction among successfully cleaned casts:

```{r example-profile}
#| fig-cap: "Example raw vs cleaned temperature profile for the cast with the largest row reduction."

example <- cast_summary %>%
  filter(status %in% c("cleaned", "skipped"), raw_rows > 0) %>%
  slice_max(pct_removed, n = 1, with_ties = FALSE)

if (nrow(example) == 1) {
  raw_file <- here("data", "01_raw", example$cruise_id, paste0(example$cast_id, ".csv"))
  clean_file <- here("data", "02_clean", paste0(example$cast_id, ".csv"))

  raw_df <- read_csv(raw_file, show_col_types = FALSE) %>%
    filter(!is.na(sea_water_pressure))
  clean_df <- read_csv(clean_file, show_col_types = FALSE)

  raw_plot <- raw_df %>%
    mutate(stage = "Raw") %>%
    select(stage, sea_water_pressure, sea_water_temperature)
  clean_plot <- clean_df %>%
    mutate(stage = "Clean") %>%
    select(stage, sea_water_pressure, sea_water_temperature)

  bind_rows(raw_plot, clean_plot) %>%
    ggplot(aes(sea_water_temperature, sea_water_pressure, color = stage)) +
    geom_path(linewidth = 0.6) +
    scale_y_reverse() +
    labs(
      title = paste0(example$cast_id, " (", example$pct_removed, "% rows removed)"),
      x = "Temperature (°C)",
      y = "Pressure (dbar)",
      color = NULL
    ) +
    theme_minimal(base_size = 12)
} else {
  cat("No cleaned casts available for profile comparison.")
}
```

## Data locations

| Stage | Path |
|-------|------|
| Raw downloads | `data/01_raw/{cruise_id}/{erddap_dataset_id}.csv` |
| Cleaned casts | `data/02_clean/{erddap_dataset_id}.csv` |
| Processing log | `data/02_clean/processing_log.csv` |

Generated on `r format(Sys.time(), "%Y-%m-%d %H:%M %Z")`.
