---
title: "Stations Across Cruises"
subtitle: "How the CTD station network has grown, shifted, and turned over"
execute:
  echo: false
  message: false
  warning: false
---

This vignette explores the **sampling design** itself, independent of any oxygen values: which stations were occupied on which cruises.

```{r setup}
ensure_packages <- function(pkgs) {
  for (pkg in pkgs) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg, repos = "https://cloud.r-project.org")
    }
  }
}

ensure_packages(c(
  "here", "dplyr", "ggplot2", "knitr", "scales", "viridisLite",
  "leaflet", "plotly", "htmlwidgets"
))

library(here)
library(dplyr)
library(ggplot2)
library(knitr)
library(scales)
library(viridisLite)
library(leaflet)
library(plotly)

source(here("R/station_cruise_summary.R"))

mapping <- load_station_cruise_mapping()
stations <- station_summary(mapping)
cruise_counts <- cruise_station_counts(mapping)
growth <- cumulative_station_growth(mapping)
presence <- station_presence_long(mapping)

n_cruises <- dplyr::n_distinct(mapping$cruise_id)
n_stations <- dplyr::n_distinct(mapping$station)
date_range <- range(mapping$cruise_date, na.rm = TRUE)
```

## Overview

| Metric | Value |
|--------|------:|
| Cruises | `r n_cruises` |
| Distinct stations ever occupied | `r n_stations` |
| Date range | `r format(date_range[1], "%Y-%m-%d")` to `r format(date_range[2], "%Y-%m-%d")` |
| Station-cruise occupations | `r nrow(mapping)` |
| Median stations per cruise | `r round(median(cruise_counts$n_stations))` |
| Stations occupied on ≥ 90% of cruises | `r sum(stations$n_cruises >= 0.9 * n_cruises)` |
| Stations occupied on only one cruise | `r sum(stations$n_cruises == 1)` |

## Where are the frequently- and rarely-visited stations?

Each marker is a station, colored and sized by how many cruises occupied it. 

```{r frequency-map}
#| fig-cap: "Station occupation frequency across all cruises."

freq_pal <- colorNumeric(palette = viridisLite::viridis(256), domain = stations$n_cruises)

leaflet(stations, width = "100%", height = 560) |>
  addProviderTiles(providers$Esri.OceanBasemap) |>
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    radius = ~rescale(n_cruises, to = c(3, 12), from = range(stations$n_cruises)),
    color = ~freq_pal(n_cruises),
    stroke = FALSE,
    fillOpacity = 0.85,
    popup = ~paste0(
      "<b>Station ", station, "</b><br>",
      "Cruises: ", n_cruises, "<br>",
      "First: ", first_cruise_date, "<br>",
      "Last: ", last_cruise_date
    )
  ) |>
  addLegend(
    position = "bottomright",
    pal = freq_pal,
    values = ~n_cruises,
    title = "Cruises"
  )
```

## How often is each station visited?

The table below highlights the most commonly sampled stations and the stations that were sampled only once.

::: {.panel-tabset}

## Core stations (≥ 90% of cruises)

```{r core-stations}
stations |>
  filter(n_cruises >= 0.9 * .env$n_cruises) |>
  select(station, n_cruises, first_cruise_date, last_cruise_date) |>
  kable(
    col.names = c("Station", "Cruises", "First cruise", "Last cruise"),
    caption = "Stations occupied on at least 90% of all cruises."
  )
```

## One-off stations

```{r oneoff-stations}
stations |>
  filter(n_cruises == 1) |>
  select(station, first_cruise_date, longitude, latitude) |>
  arrange(first_cruise_date) |>
  kable(
    col.names = c("Station", "Cruise date", "Longitude", "Latitude"),
    digits = 3,
    caption = "Stations occupied on exactly one cruise."
  )
```

:::

## Presence table: which stations, which cruises

Every cell is one station-cruise pair: filled if that station was occupied on that cruise. Stations (rows) are ordered by the cruise on which they **first** appear, so new stations entering the program show up as a new row starting partway across the plot — a staircase pattern reveals when the network expanded. Cruises (columns) run left to right in chronological order. Zoom and hover to identify a cell.

```{r presence-heatmap}
#| fig-cap: "Station occupation by cruise (chronological)."
#| fig-height: 16

presence_plot <- presence |>
  mutate(z = as.integer(present))

fig <- plot_ly(
  data = presence_plot,
  x = ~cruise_id,
  y = ~station,
  z = ~z,
  type = "heatmap",
  colorscale = list(c(0, "#f4f4f4"), c(1, "#21918c")),
  showscale = FALSE,
  hovertemplate = paste0(
    "Station: %{y}<br>Cruise: %{x}<br>Date: %{customdata}<extra></extra>"
  ),
  customdata = ~format(cruise_date, "%Y-%m-%d")
) |>
  layout(
    xaxis = list(title = "Cruise", tickangle = -90, tickfont = list(size = 8)),
    yaxis = list(title = "Station", tickfont = list(size = 7)),
    margin = list(b = 120)
  )

fig
```

## Stations sampled per cruise

The number of distinct stations occupied fluctuates with survey design and vessel time.

```{r stations-per-cruise}
#| fig-cap: "Distinct stations occupied, by cruise date."

ggplot(cruise_counts, aes(x = cruise_date, y = n_stations)) +
  geom_line(color = "#21918c", linewidth = 0.85) +
  geom_point(color = "#21918c", size = 1.8) +
  scale_y_continuous(name = "Stations occupied", limits = c(0, NA)) +
  labs(x = "Cruise date") +
  theme_minimal(base_size = 12)
```

## Growth of the station network

Cumulative count of distinct stations ever visited, through each cruise. Plateaus mean a cruise visited only stations that had already been sampled before; steps up mean new stations entered the program.

```{r cumulative-growth}
#| fig-cap: "Cumulative distinct stations occupied, over time."

ggplot(growth, aes(x = cruise_date, y = cumulative_stations)) +
  geom_step(color = "#3b528b", linewidth = 0.9) +
  geom_point(color = "#3b528b", size = 1.4) +
  scale_y_continuous(name = "Cumulative distinct stations") +
  labs(x = "Cruise date") +
  theme_minimal(base_size = 12)
```
