---
title: "SV18067 Report"
format: html
params:
  batch_value: "SV18067"
  batch_name: "cruise_anoxic_depth"
---

```{r setup}
#| include: false
ensure_packages <- function(pkgs) {
  for (pkg in pkgs) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg, repos = "https://cloud.r-project.org")
    }
  }
}

ensure_packages(c(
  "here", "glue", "dplyr", "leaflet",
  "viridisLite", "htmlwidgets", "jsonlite"
))

library(here)
library(glue)
library(dplyr)
library(leaflet)
library(viridisLite)
library(htmlwidgets)
library(jsonlite)
```

```{r get-data}
#| code-summary: Load cleaned observations and interpolated oxygen field
source(here(glue("{params$batch_name}/getData.R")))
source(here(glue("{params$batch_name}/anoxic_depth.R")))
source(here(glue("{params$batch_name}/plot_anoxic_depth_map.R")))

data <- getData(params$batch_value)
obs <- data$observations
field <- data$oxygen_field
```

## Cruise `r params$batch_value`

This report maps the **shallowest depth** at which the interpolated dissolved oxygen field falls below a chosen threshold (anoxic/hypoxic boundary). Use the slider on the map to change the O₂ threshold; grid cells are colored by **depth in meters**.

```{r data-summary}
#| code-summary: Observation and field summary
data.frame(
  casts = length(unique(obs$cast_id)),
  observation_rows = nrow(obs),
  field_grid_points = if (is.null(field)) NA_integer_ else nrow(field),
  oxygen_range_mg_l = if ("dissolved_oxygen" %in% names(obs)) {
    range(obs$dissolved_oxygen, na.rm = TRUE)
  } else {
    c(NA_real_, NA_real_)
  }
)
```

## Anoxic depth map

Pan and zoom to explore where oxygen drops below the threshold. Each colored cell shows the **shallowest depth** (closest to the surface) where interpolated O₂ is below the threshold — not the depth where O₂ equals the threshold. When the threshold exceeds surface oxygen across the domain, expect uniform shallow depths (~2 m, the top of the interpolated grid). Hover a cell for its depth in meters. The color bar uses a **fixed depth scale** so colors are comparable across thresholds.

```{r anoxic-depth-map}
#| code-summary: Interactive map of shallowest anoxic depth

if (is.null(field)) {
  cat(
    "Interpolated field not found. Run `make interpolate CRUISE=",
    params$batch_value,
    "` first.",
    sep = ""
  )
} else {
  map <- plot_anoxic_depth_map(field, obs)
  if (is.null(map)) {
    cat("No interpolated field grid available for this cruise.")
  } else {
    map
  }
}
```
