---
title: "SV18067 Report"
format: html
params:
  batch_value: "SV18067"
  batch_name: "cruise_oxygen"
  map_depth_m: null
---

```{r setup}
#| include: false
ensure_packages <- function(pkgs) {
  for (pkg in pkgs) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg, repos = "https://cloud.r-project.org")
    }
  }
}

ensure_packages(c(
  "here", "glue", "dplyr", "leaflet",
  "viridisLite", "htmlwidgets", "jsonlite"
))

library(here)
library(glue)
library(dplyr)
library(leaflet)
library(viridisLite)
library(htmlwidgets)
library(jsonlite)
```

```{r get-data}
#| code-summary: Load cleaned observations and interpolated oxygen field
source(here(glue("{params$batch_name}/getData.R")))
source(here(glue("{params$batch_name}/depth_layers.R")))
source(here(glue("{params$batch_name}/plot_oxygen_map.R")))

data <- getData(params$batch_value)
obs <- data$observations
field <- data$oxygen_field
```

## Cruise `r params$batch_value`

This report compares the DIVAnd-interpolated dissolved oxygen field with cleaned CTD observations from `data/02_clean/`.

```{r data-summary}
#| code-summary: Observation and field summary
data.frame(
  casts = length(unique(obs$cast_id)),
  observation_rows = nrow(obs),
  field_grid_points = if (is.null(field)) NA_integer_ else nrow(field),
  oxygen_range_mg_l = if ("dissolved_oxygen" %in% names(obs)) {
    range(obs$dissolved_oxygen, na.rm = TRUE)
  } else {
    c(NA_real_, NA_real_)
  }
)
```

## Interpolated field map

The map below is interactive: pan and zoom to explore the gridded dissolved oxygen field. Use **Prev** / **Next** to step through depth bands. CTD observations are colored with the same palette as the interpolated field.

```{r oxygen-map}
#| code-summary: Interactive map with interpolated field and observation points

if (is.null(field)) {
  cat("Interpolated field not found. Run `make interpolate CRUISE=", params$batch_value, "` first.", sep = "")
} else {
  map <- plot_oxygen_map(field, obs)
  if (is.null(map)) {
    cat("No field data available near the selected depth.")
  } else {
    map
  }
}
```
